---
title: "nf-g4rnaseq report"
subtitle: "RNA-seq differential expression + G-quadruplex / RNA:DNA hybrid-G4 analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({ library(data.table); library(knitr) })
ex  <- function(p) file.exists(p)
img <- function(p, cap="") if (ex(p)) cat(sprintf("\n![%s](%s)\n\n", cap, p))
rd  <- function(p) if (ex(p)) fread(p) else NULL
# result directories are staged with these names (placeholders if a stage was skipped)
```

This report is generated automatically from the pipeline results. Sections appear only for stages that were run. See `docs/output.md` for column definitions.

```{r de, results='asis'}
if (dir.exists("deseq2")) {
  cat("# Differential expression\n\n")
  s <- rd("deseq2/deg_summary.csv")
  if (!is.null(s)) { cat("Significant genes per contrast (padj threshold):\n\n"); print(kable(s)); cat("\n\n") }
  cat("## Quality control\n\n")
  img("deseq2/figures/PCA.png", "PCA of variance-stabilised counts")
  img("deseq2/figures/sample_distance_heatmap.png", "Sample-to-sample distance")
  img("deseq2/figures/dispersion.png", "Dispersion estimates")
  cat("## Per-contrast results\n\n")
  csvs <- setdiff(list.files("deseq2", pattern="\\.csv$"), "deg_summary.csv")
  for (cf in csvs) {
    id <- sub("\\.csv$","",cf); d <- rd(file.path("deseq2", cf))
    if (is.null(d)) next
    cat(sprintf("### %s\n\n", id))
    nm <- if ("std_name" %in% names(d)) "std_name" else "gene_id"
    top <- head(d[order(d$padj)][, intersect(c(nm,"log2FoldChange","padj","baseMean"), names(d)), with=FALSE], 8)
    print(kable(top, digits=3, caption=sprintf("Top genes by padj — %s", id)))
    img(sprintf("deseq2/figures/volcano_%s.png", id))
    cat("\n\n")
  }
}
```

```{r g4, results='asis'}
if (dir.exists("g4") && ex("g4/g4_sites.tsv")) {
  g <- fread("g4/g4_sites.tsv")
  cat("# G-quadruplex landscape\n\n")
  cat(sprintf("- **%d** predicted G4 sites (+ strand: %d, − strand: %d).\n", nrow(g), sum(g$strand=="+"), sum(g$strand=="-")))
  cat(sprintf("- canonical-strength (≥3 tetrads): %d; two-quartet: %d; high-confidence (score≥47): %d.\n\n",
              sum(g$n_tetrads>=3), sum(g$n_tetrads==2), sum(g$score>=47)))
  f <- rd("g4/gene_g4_features.tsv")
  if (!is.null(f)) {
    cat(sprintf("- **%.1f%%** of genes have a proximal G4 (any, permissive); **%.1f%%** high-confidence.\n\n",
                100*mean(f$has_g4), 100*mean(f$has_g4_highconf)))
    cat("> Permissive G4 calls cover most genes and have low specificity — base any enrichment *claim* on the high-confidence subset.\n\n")
  }
}
```

```{r hybrid, results='asis'}
hf <- if (ex("hybrid/gene_hybrid_g4.tsv")) "hybrid/gene_hybrid_g4.tsv" else if (ex("g4/gene_hybrid_g4.tsv")) "g4/gene_hybrid_g4.tsv" else NA
if (!is.na(hf)) {
  h <- fread(hf)
  cat("# RNA:DNA hybrid G-quadruplex (PHQS)\n\n")
  cat("Implements the co-transcriptional hybrid-G4 model (Zheng et al. 2013): a gene whose coding strand has only 2–3 G-tracts — too few for a DNA-only G4 — can still form a G4 with the nascent transcript.\n\n")
  cat(sprintf("- hybrid-**capable** (≥%s tracts): **%.1f%%** of genes; hybrid-**specific** (2–3 tracts, no DNA-only G4): **%.1f%%**.\n\n",
              "2", 100*mean(h$hybrid_capable), 100*mean(h$hybrid_specific)))
  hs <- head(h[hybrid_specific==TRUE][order(sense_G_frac)][, intersect(c("std_name","biotype","sense_G_frac","sense_G3_tracts"), names(h)), with=FALSE], 8)
  if (nrow(hs)) print(kable(hs, caption="Lowest-guanine hybrid-specific genes"))
  cat("\n> Hybrid-G4 potential is a *sequence prediction* (not an in-vivo measurement) and scales with gene length; treat expression associations as length-confounded and hypothesis-generating, to be confirmed orthogonally (e.g. BG4/DRIP).\n\n")
}
```

```{r g4enrich, results='asis'}
if (ex("g4enrich/g4_deg_enrichment.csv")) {
  e <- fread("g4enrich/g4_deg_enrichment.csv")
  cat("# Are differentially-expressed genes enriched for G4s?\n\n")
  key <- e[get("set")=="with_ribo" & direction=="all" & flag %in% c("has_g4_highconf","hybrid_capable")]
  if (nrow(key)) print(kable(key[, intersect(c("contrast","flag","pct_deg","pct_bg","odds_ratio","p"), names(key)), with=FALSE], digits=3))
  cat("\n\n")
}
```

```{r go, results='asis'}
if (dir.exists("go")) {
  gos <- list.files("go", pattern="^GO_BP_.*csv$")
  if (length(gos)) {
    cat("# GO enrichment (top biological process per contrast)\n\n")
    for (gf in gos) { d <- rd(file.path("go", gf)); if (!is.null(d) && nrow(d))
      cat(sprintf("- **%s**: %s (p.adj %.1g)\n", sub("GO_BP_|\\.csv","",gf), d$Description[1], d$p.adjust[1])) }
    cat("\n")
    img(list.files("go/figures", pattern="dotplot_GOBP.*down.*png", full.names=TRUE)[1])
  }
}
```

```{r gsea, results='asis'}
if (dir.exists("gsea")) {
  cat("# Gene-set analysis\n\n")
  gs <- list.files("gsea", pattern="^g4_geneset_camera_.*csv$")
  if (length(gs)) cat("**G4 gene sets** — competitive, correlation-aware test (`limma::cameraPR`; BH-FDR). A negative median statistic = the set is down-regulated in that contrast:\n\n")
  for (gf in gs) { d <- rd(file.path("gsea", gf)); if (!is.null(d))
    cat(sprintf("- **%s**: %s\n", sub("g4_geneset_camera_|\\.csv","",gf),
        paste(sprintf("%s %s (median stat %.2f, FDR %.2g)", d$ID, d$Direction, d$median_stat, d$padj), collapse="; "))) }
  cat("\n\n"); img("gsea/figures/G4_geneset_camera_byContrast.png", "G4 gene sets across contrasts (median Wald statistic; cameraPR BH-FDR stars)")
  if (length(list.files("gsea", pattern="^gseGO_"))) cat("\nGO biological-process GSEA (focused sets, fgsea) tables: `gsea/gseGO_*.csv`.\n\n")
}
```

```{r integrate, results='asis'}
if (ex("integrate/master_gene_table.tsv")) {
  m <- fread("integrate/master_gene_table.tsv")
  cat(sprintf("# Integrated table\n\nThe master per-gene table (`integrate/master_gene_table.tsv`) joins expression, per-contrast statistics, proximal-G4 features and hybrid-G4 class for **%d** genes — the backbone for follow-up.\n\n", nrow(m)))
}
```

# Methods & caveats

Reads were quantified per gene (featureCounts; strandedness auto-detected) and modelled with **DESeq2** using the configured design; fold-changes are shrunken (apeglm/ashr). G4s were predicted with **pqsfinder** (both strands); the RNA:DNA hybrid-G4 (PHQS) scan counts clustered coding-strand G-tracts. Functional enrichment used **clusterProfiler** (GO/KEGG over-representation + GO-set GSEA). For the large G4 gene sets we use **limma::cameraPR** — a competitive, correlation-aware test calibrated for any set size — rather than fgsea, whose p-values inflate with set size. G4↔DE Fisher tests are **BH-corrected** across the family. Interpret with attention to: replication/power (the pipeline warns at <3 replicates per group; gene-set tests are more robust than per-gene tests at low n), the low specificity of permissive G4 calling (prefer the high-confidence subset), the predictive/length-confounded nature of the hybrid-G4 calls, and any confounded design factors (flag via `--flag_genes`). Full tool versions and citations: `CITATIONS.md`.
